Windowed least-squares model reduction for dynamical systems

### 2019-10-08 section4/numerical_methods.tex
through space--time, proofing

\section{Numerical solution techniques}\label{sec:numerical_techniques}
The discussion to this point has outlined the formulation of the \methodAcronym\ method. We now consider numerical solution techniques. 
\EP{I think passive voice is fine in this instance, it reads easier}
As \methodAcronym\ displays commonalities with optimal
control problems, a large existing body of work can be leveraged. This section
discuses a variety of such techniques and outlines how they can be used to
solve the \methodAcronym\ system. We focus primarily on 
%\KTC{earlier we kept
%saying 
%direct (i.e.,
%	discretize-then-optimize) and indirect (i.e., optimize-then-discretize).
%	Let's be consistent when we re-introduce it here.}
direct (discretize-the-optimize) and indirect (optimize-then-discretize) for \spatialAcronym\ trial subspaces, 
and then briefly outline direct and indirect methods for \spaceTimeAcronym\ trial subspaces.

\subsection{\spatialAcronym: direct and indirect methods}
Numerical techniques to
solve optimal control problems can be classified as either
\textit{direct} or \textit{indirect}
methods~\cite{conway_optimalcontrolreview}.  Direct methods eschew the first order optimality conditions
~\eqref{eq:lspg_continuous}--\eqref{eq:lspg_bcs} and, instead, ``directly" solve the optimization problem~\eqref{eq:tclsrm} by first
numerically discretizing the system and objective functional and ``transcribing"
the infinite dimensional problem into a finite-dimensional nonlinear
programming (optimization) problem. Direct approaches thus ``discretize then optimize."
Indirect methods, on the other hand, leverage the calculus of variations to
derive the Euler--Lagrange equations, which comprise the first-order
optimality conditions~\eqref{eq:lspg_continuous}--\eqref{eq:lspg_bcs} associated with the objective 
functional. Indirect methods then solve (numerically) the Euler--Lagrange equations. Thus, indirect methods ``optimize then discretize," and 
solve the optimization problem
``indirectly". A variety of discretization/solution techniques are possible for both direct and indirect methods. Collocation methods,
finite elements, spectral methods, shooting methods, etc., are all examples of
plausible solution techniques.

In the present context, we investigate both direct and indirect methods to solve \methodAcronym\ with \spatialAcronym\ trial
subspaces. We consider: 
\begin{itemize} \item \textit{Direct Methods (discretize-then-optimize)}: Discretize and then optimize approaches are outlined for linear multistep methods. 
\item \textit{Indirect methods (optimize-then-discretize)}: An indirect method leveraging the forward-backward sweep algorithm is outlined 
to solve~\eqref{eq:lspg_continuous}-\eqref{eq:lspg_bcs}. 
\end{itemize} 
We emphasize that a variety of other approaches exist. Future work will comprise a detailed exploration
of additional approaches.
Section~\ref{sec:direct}  outlines direct solution approaches based on linear
multistep methods, while Section~\ref{sec:indirect} outlines the indirect
solution approach.

>>>>

\section{Numerical solution techniques}\label{sec:numerical_techniques}
The discussion to this point has outlined the formulation of the \methodAcronym\ method. We now consider numerical solution techniques. 
\methodAcronym\ displays commonalities with optimal
control problems, and as such a large existing body of work can be leveraged. In this section we 
outline a variety of such techniques. We focus primarily on 
%\KTC{earlier we kept
%saying 
%direct (i.e.,
%	discretize-then-optimize) and indirect (i.e., optimize-then-discretize).
%	Let's be consistent when we re-introduce it here.}
direct (discretize-the-optimize) and indirect (optimize-then-discretize) for \spatialAcronym\ trial subspaces, 
and then briefly outline direct and indirect methods for \spaceTimeAcronym\ trial subspaces.

\subsection{\spatialAcronym: direct and indirect methods}
Numerical techniques to
solve optimal control problems can be classified as either
\textit{direct} or \textit{indirect}
methods~\cite{conway_optimalcontrolreview}. Rather than working with the first order optimality conditions~\eqref{eq:lspg_continuous}--\eqref{eq:lspg_bcs}, direct methods ``directly" solve the optimization problem~\eqref{eq:tclsrm} by first
numerically discretizing the system and objective functional and ``transcribing"
the infinite dimensional problem into a finite-dimensional nonlinear
programming (optimization) problem. Direct approaches thus ``discretize-then-optimize."
Indirect methods, on the other hand, compute solutions to the Euler--Lagrange equations~\eqref{eq:lspg_continuous}--\eqref{eq:lspg_bcs} (which comprise the first order optimality conditions of the objective functional). Thus, indirect methods ``optimize-then-discretize," and solve the optimization problem
``indirectly". A variety of discretization/solution techniques are possible for both direct and indirect methods. Collocation methods,
finite elements, spectral methods, shooting methods, etc., are all examples of
plausible solution techniques.

In the present context, we investigate both direct and indirect methods to solve \methodAcronym\ with \spatialAcronym\ trial
subspaces. We consider: 
\begin{itemize} \item \textit{Direct methods (discretize-then-optimize)}: A direct approach leveraging linear multistep methods to directly solve the optimization problem~\eqref{eq:tclsrm}.
\item \textit{Indirect methods (optimize-then-discretize)}: An indirect approach leveraging the forward-backward sweep algorithm to solve the Euler--Lagrange equations~\eqref{eq:lspg_continuous}-\eqref{eq:lspg_bcs}.

\end{itemize} 
We emphasize that a variety of other approaches exist. Future work will comprise a detailed exploration
of additional approaches.
Section~\ref{sec:direct}  outlines direct solution approaches based on linear
multistep methods, while Section~\ref{sec:indirect} outlines the indirect
solution approach.

### 2019-10-08 section4/spacetime_numerical.tex
through space--time, proofing

\subsection{\spaceTimeAcronym\ trial spaces: direct solution technique}
The direct solution technique seeks to minimize the fully discrete objective function associated with the O$\Delta$E. For brevity, we again focus on linear multistep methods.

@@

Leveraging the definition of the fully discrete objective function~\eqref{eq:obj_lms}, and assuming $N \nstepsArg{n} \ge \stdimArg{n}$, \methodAcronym\ leveraging a \spaceTimeAcronym\ basis is given by, 
 \begin{equation}\label{eq:obj_gen_lms_final_st}
\begin{split}
& \underset{\stgenstatey}{\text{minimize } }
\objectiveArgLMS{n} (\stbasisArg{n}{\timeWindowArg{n}{0}} \stgenstatey + \stateInterceptArg{n},\ldots,\stbasisArg{n}{\timeWindowArg{n}{\nstepsArg{n}}} \stgenstatey + \stateInterceptArg{n})  \\
&\text{subject to } \;  \stbasisArg{n}{\timeStartArg{n}}\stgenstateyArg{n} = \begin{cases} 
\mathbb{P}^n(\stbasisArgnt{n-1}\stgenstateArg{n-1})(\timeEndArg{n-1}) & n = 2,\ldots,\nslabs, \\ 
\mathbb{P}^n(\stateFOMArgnt{1})(0) &
n=1. \end{cases}\end{split} 
\end{equation}

% \begin{equation}\label{eq:obj_gen_lms_final_st}
%\begin{split}
%&\stgenstateArg{n} = \underset{\stgenstatey}{\text{arg\,min } }
%\objectiveArgLMS{n} (\stbasisArg{n}{\timeWindowArg{n}{0}} \stgenstatey + \stateInterceptArg{n},\ldots,\stbasisArg{n}{\timeWindowArg{n}{\nstepsArg{n}}} \stgenstatey + \stateInterceptArg{n})  \\
%&\text{subject to } \;  \stbasisArg{n}{\timeStartArg{n}}\stgenstateArg{n} = \begin{cases} 
%\mathbb{P}^n(\stbasisArgnt{n-1}\stgenstateArg{n-1})(\timeEndArg{n-1}) & n = 2,\ldots,\nslabs, \\ 
%\mathbb{P}^n(\stateFOMArgnt{1})(0) &
%n=1. \end{cases}\end{split} 
%\end{equation}

The minimization problem~\eqref{eq:obj_gen_lms_final_st} can be solved, e.g, via the Gauss-Newton method.

\begin{remark}
For the limiting case where the window size spans the entire domain, $\DeltaSlabArg{1}\equiv T$, direct \methodAcronym\ with a \spaceTimeAcronym\ trial subspace recovers ST-LSPG.
\end{remark}

\subsection{\spaceTimeAcronym\ trial spaces: indirect solution technique}
As opposed to the direct solution technique, the indirect solution technique directly minimizes the continuous objective function~\eqref{eq:obj_gen_slab_spacetime},

@@

 & \underset{\stgenstatey}{\text{minimize }} \mathcal{J}^n \bigg( \stbasisArgnt{n} \stgenstatey + \stateInterceptArg{n}  \bigg) \\  
&\text{subject to } \;  \stbasisArg{n}{\timeStartArg{n}}\stgenstateyArg{n} = \begin{cases} 
\mathbb{P}^n(\stbasisArgnt{n-1}\stgenstateArg{n-1})(\timeEndArg{n-1}) & n = 2,\ldots,\nslabs, \\ 
\mathbb{P}^n(\stateFOMArgnt{1})(0) &
n=1. \end{cases} 

@@

discretization of the integral. We introduce $\ncollocSTArg{n} \ge \text{ciel}(\stdim / N )$ collocation points over the $n$th window, 
$$\timeStartArg{n} \le \collocPointSTArg{n}{1} \le \ldots \le \collocPointSTArg{n}{\ncollocSTArg{n}} \le \timeEndArg{n}.$$

@@

&\underset{\stgenstatey}{\text{minimize }} \objectiveDiscreteSTArg{n} \bigg( \stbasisArgnt{n} \stgenstatey + \stateInterceptArg{n}  \bigg) , \\
&\text{subject to }\;  \stbasisArg{n}{\timeStartArg{n}}\stgenstateyArg{n} = \begin{cases} 
\mathbb{P}^n(\stbasisArgnt{n-1}\stgenstateArg{n-1})(\timeEndArg{n-1}) & n = 2,\ldots,\nslabs, \\ 
\mathbb{P}^n(\stateFOMArgnt{1})(0) &
n=1,
\end{cases}

@@

\bigg[ \stbasisDotArg{n}{\collocPointSTArg{n}{i}} \stgenstateyArg{ }  - \velocity (\stbasisArg{n}{\collocPointSTArg{n}{i}} \stgenstateyArg{ } + \stateInterceptArg{n},\collocPointSTArg{n,i}) \bigg]^T 

@@

\bigg[ \stbasisDotArg{n}{\collocPointSTArg{n}{i}} \stgenstateyArg{ }  - \velocity (\stbasisArg{n}{\collocPointSTArg{n}{i}} \stgenstateyArg{ } + \stateInterceptArg{n},\collocPointSTArg{n,i}) \bigg].

@@

The nonlinear least-squares problem can again be solved, e.g., via the Gauss-Newton method. 

@@

\subsection{\spaceTimeAcronym\ trial space: summary}
As was just shown, \spaceTimeAcronym\ trial subspaces yield a series of space--time algebraic systems over each window. As a variety of work has examined space--time reduced-order models with \spaceTimeAcronym\ trial subspaces, 
a detailed exposition of solution techniques for these systems is not pursued here. It is sufficient to say that the 
space--time trial subspace yields a series of dense (non)linear systems to be solved over each window.

>>>>

\subsection{\spaceTimeAcronym: direct solution approach}
The direct solution technique seeks to minimize the fully discrete objective function associated with the O$\Delta$E. For brevity, we again focus on linear multistep methods.

@@

Leveraging the definition of the fully discrete objective function~\eqref{eq:obj_lms}, and assuming $N \nstepsArg{n} \ge \stdimArg{n}$, \methodAcronym\ leveraging a \spaceTimeAcronym\ basis sequentially computes the generalized coordinates $\stgenstateArg{n}$, $n=1,\ldots,\nslabs$, that satisfy
 \begin{equation}\label{eq:obj_gen_lms_final_st}
\begin{split}
& \underset{\stgenstatey}{\text{minimize } }
\objectiveArgLMS{n} (\stbasisArg{n}{\timeWindowArg{n}{0}} \stgenstatey + \stateInterceptSTArg{n},\ldots,\stbasisArg{n}{\timeWindowArg{n}{\nstepsArg{n}}} \stgenstatey + \stateInterceptSTArg{n}).%  \\
%&\text{subject to } \;  \stbasisArg{n}{\timeStartArg{n}}\stgenstateyArg{n} = \begin{cases} 
%\mathbb{P}^n(\stbasisArgnt{n-1}\stgenstateArg{n-1})(\timeEndArg{n-1}) & n = 2,\ldots,\nslabs, \\ 
%\mathbb{P}^n(\stateFOMArgnt{1})(0) &
%n=1. \end{cases}
\end{split} 
\end{equation}
% \begin{equation}\label{eq:obj_gen_lms_final_st}
%\begin{split}
%&\stgenstateArg{n} = \underset{\stgenstatey}{\text{arg\,min } }
%\objectiveArgLMS{n} (\stbasisArg{n}{\timeWindowArg{n}{0}} \stgenstatey + \stateInterceptArg{n},\ldots,\stbasisArg{n}{\timeWindowArg{n}{\nstepsArg{n}}} \stgenstatey + \stateInterceptArg{n})  \\
%&\text{subject to } \;  \stbasisArg{n}{\timeStartArg{n}}\stgenstateArg{n} = \begin{cases} 
%\mathbb{P}^n(\stbasisArgnt{n-1}\stgenstateArg{n-1})(\timeEndArg{n-1}) & n = 2,\ldots,\nslabs, \\ 
%\mathbb{P}^n(\stateFOMArgnt{1})(0) &
%n=1. \end{cases}\end{split} 
%\end{equation}
It is noted that the boundary conditions are automatically satisfied through the definition of the $\spaceTimeAcronym$\ trial subspace. The minimization problem~\eqref{eq:obj_gen_lms_final_st} can be solved, e.g, via the Gauss-Newton method.

\begin{remark}
For the limiting case where the window size spans the entire domain, $\DeltaSlabArg{1}\equiv T$, and a uniform quadrature weights are used, direct \methodAcronym\ with a \spaceTimeAcronym\ trial subspace recovers ST-LSPG.
\end{remark}

\subsection{\spaceTimeAcronym: indirect solution approach}
As opposed to the direct approach, the indirect approach directly minimizes the continuous objective function~\eqref{eq:obj_gen_slab_spacetime},

@@

 & \underset{\stgenstatey}{\text{minimize }} \mathcal{J}^n \bigg( \stbasisArgnt{n} \stgenstatey + \stateInterceptSTArg{n} \otimes \onesFunctionArg{n} \bigg) .%\\  
%&\text{subject to } \;  \stbasisArg{n}{\timeStartArg{n}}\stgenstateyArg{n} = \begin{cases} 

@@

discretization of the integral. We introduce $\ncollocSTArg{n} \ge \text{ceil}(\stdimArg{n}/ N )$ collocation points over the $n$th window, 
$$\timeStartArg{n} \le \collocPointSTArg{n}{1} \le \cdots \le \collocPointSTArg{n}{\ncollocSTArg{n}} \le \timeEndArg{n}.$$

@@

&\underset{\stgenstatey}{\text{minimize }} \objectiveDiscreteSTArg{n} \bigg( \stbasisArgnt{n} \stgenstatey + \stateInterceptSTArg{n} \otimes \onesFunctionArg{n}  \bigg) , %\\
%&\text{subject to }\;  \stbasisArg{n}{\timeStartArg{n}}\stgenstateyArg{n} = \begin{cases} 

@@

\bigg[ \stbasisDotArg{n}{\collocPointSTArg{n}{i}} \stgenstateyArg{ }  - \velocity (\stbasisArg{n}{\collocPointSTArg{n}{i}} \stgenstateyArg{ } + \stateInterceptSTArg{n},\collocPointSTArg{n,i}) \bigg]^T 

@@

\bigg[ \stbasisDotArg{n}{\collocPointSTArg{n}{i}} \stgenstateyArg{ }  - \velocity (\stbasisArg{n}{\collocPointSTArg{n}{i}} \stgenstateyArg{ } + \stateInterceptSTArg{n},\collocPointSTArg{n,i}) \bigg].

@@

The least-squares problem can again be solved, e.g., via the Gauss-Newton method. 

@@

\subsection{\spaceTimeAcronym\: summary}
\spaceTimeAcronym\ trial subspaces yield a series of space--time algebraic systems over each window. As a variety of work has examined space--time reduced-order models with \spaceTimeAcronym\ trial subspaces, 
a detailed exposition of solution techniques for these systems is not pursued here. It is sufficient to say that the 
space--time trial subspace yields a series of dense (non)linear systems to be solved over each window.
